Unsupervised Learning for Identifying Events in Active Target Experiments

### 2020-08-05 doc/Manuscript/article.tex
Updates from Overleaf

\begin{abstract}
% Modified by DB (6/8/2020)
This article presents novel applications of machine learning methods to the problem of event separation in an active target detector, the Active-Target Time Projection Chamber (AT-TPC) \cite{Bradt2017}. 
The overarching goal is to identify classes of events in the early stages of the data analysis, thereby improving its efficiency by limiting the computationally expensive processing to events of interest only, which are often not the most abundant.
%Machine learning presents an interesting avenue for researchers operating the AT-TPC, as traditional analysis methods of AT-TPC data are computationally expensive and have to fit all tracks against the event type of interest. The latter presents a considerable challenge when either the space of reactions is not known prior to the analysis, or the event type of interest is not the most abundant. 
The application of unsupervised clustering  algorithms to the analysis of two-dimensional projections of particle tracks from a resonant proton scattering experiment on $^{46}$Ar is introduced. We explore the performance of autoencoder neural networks and a pre-trained VGG16 \cite{Simonyan2014} convolutional neural network.
%on two tasks: a semi-supervised classification task and the unsupervised clustering of particle tracks. 
%On the semi-supervised task, we find that a logistic regression classifier trained on small labeled subsets of the latent space of these models perform very well, both on simulated and real data. On the clustering task, 
We find that a $K$-means algorithm applied to the simulated data in the VGG16 latent space forms almost perfect clusters. Additionally, the VGG16+K-means approach finds high purity clusters of proton events for real experimental data. We explore also the application of autoencoder neural networks to clustering by implementing. With autoencoder neural networks we find improved descriptions of data from experiments.

@@

The AT-TPC \cite{Bradt2017} is a novel type of detector designed specifically for nuclear physics experiments where the energies of the recoiling particles are very low compared to the energy required to escape target material. The luminosity of nuclear physics experiments performed with fixed targets is directly proportional to the amount of material encountered by the beam. On the other hand, for several classes of experiments the detection of recoil particles is paramount, therefore limiting the target thickness. In addition, the properties of the recoil particles are modified while traversing the target material, affecting the resolutions that can be achieved. This necessary balance between luminosity and resolution is particularly difficult when performing experiments with rare isotope beams, because of the low intensities available.

The concept of active targets aim at mitigating this compromise, by turning the target itself into a detector \cite{BECEIRONOVO2015}. Most active target detectors such as the AT-TPC are composed of a time projection chamber (TPC) where the detector gas is at the same time the target material. Recoil particles that originate from a nuclear reaction between a beam nucleus and a gas nucleus can be tracked from the vertex of the reaction to their final position inside the active volume of the target. Their properties can therefore be measured without any loss of resolution regardless of the amount of material traversed by the beam. At the same time, the detection efficiency is dramatically increased by the very large solid angle covered in this geometry. A direct consequence of this concept is the inclusiveness of the experimental data recorded by this type of detector: any nuclear reaction happening within the target is recorded. Although this sounds like an advantage from the scientific point of view, it poses great challenges during the analysis phase, that are reminiscent of bubble chamber times and on par with event classification challenges in particle physics today, see for example the recent review of Mehta {\em et al.} \cite{mehta2019}. More often than not, the reaction channel of interest has one of the lowest cross sections, therefore the physicist is faced with the task of sorting out the corresponding events from the ``background" of other reaction channels.

% motivating ML work:
Because TPCs produce three-dimensional images of charged particle tracks, the event identification task is often akin to a visual inspection (comparable to the analyses made in the bubble chamber era), which is not practical nowadays because of the large quantities of data. Machine learning techniques then appear as a promising prospect, in particular in the image recognition domain where much progress has been made recently \cite{mehta2019}. In addition, machine learning (ML) algorithms offer new possibilities such as the potential discovery of unforeseen phenomena that would have been missed by more traditional analysis methods. Prior work has demonstrated the ability to apply supervised classification machine learning methods to AT-TPC data when a labeled training set is available, whether through hand-labeled data or labeled simulated data (a {\em transfer learning} application) \cite{Kuchera2019}. In some experiments, a labeled data set is unavailable. This could be  due to the inability to hand-label events, or the case where one does not know {\em a priori} the types and behaviors of the reactions present in the detector in order to generate a labeled, simulated data set. In the latter case, there must also exist a validation data set of real data, still requiring the ability to label a subset of the real data. The unsupervised separation of event types, or {\em clustering}, based on a set of ML algorithms is hereby examined, using experimental data recorded by the AT-TPC during its commissioning experiment from a radioactive $^{46}$Ar beam reacting on an isobutane target composed of proton and carbon nuclei.

@@

The detector was placed inside the bore of a MRI solenoid energized to $\sim 2$ Tesla. This axial magnetic field served the purpose of bending the trajectories of the recoil particles in order to i) increase their length and ii) provide a measurement of their bending radius, directly related to their magnetic rigidity. Because the recoil particles travel in gas, they slow down and eventually stop, therefore their trajectories are described by three-dimensional spirals (see \cite{Bradt2017}). One of the difficulties encountered in the analysis is that the shape of these spirals does not have an analytical form because it follows the energy-loss profile of the particles. It therefore needs to be simulated via an integration, which is numerically costly. Other difficulties are related to several experimental effects that deteriorate the quality of the data, namely saturation and cross-talk effects, as well as random noise.

The method used in \cite{Bradt2018} to analyze the data followed a three-phase sequence: cleaning, filtering and fitting. Traditional methods were used to perform each of these tasks, and ultimately extract the scientific information, but there were severe limitations and high computational costs that become prohibitive in data sets larger by an order of magnitude.  In particular, as for the data presented and analyzed here, with data sets in the terabyte region or larger, these more traditional methods become prohibitively expensive from a computational stand.
% Begin added by DB (06/10/2020)
The cleaning was performed using a combination of linear and circular Hough transforms on a 2-dimensional projection of the tracks \cite{Bradt2017}. The following filtering and fitting phases were performed simultaneously, by defining the cost function to the fitting algorithm as a sum of three $\chi^2$ components based on i) the position of the track in space, ii) the energy deposited on each pixel of the sensor plane, and iii) the location of the vertex of the reaction. While various fitting algorithms were tested, the most accurate was a Monte-Carlo algorithm that explored the six-dimensional phase space of the particle's kinematics parameters, reducing it progressively at each iteration step until the desired accuracy was reached \cite{Bradt2017}. Although this algorithm ended up being the most accurate, it is extremely costly computationally because of the very large number of simulated tracks needed for each event. The filtering was performed by setting limits to the $\chi^2$ distributions, below which the events were assigned as proton scattering. This is a very inefficient method because it requires performing the fitting for all events, including those that are not of interest. Pioneering work on event identification using machine learning methods, namely the use of a pre-trained convolutional neural networks (CNN), later showed that the filtering phase would better be performed using this type of technique \cite{Kuchera2019}. 
% End added by DB (06/10/2020)
In addition, the authors of Ref.~\cite{Kuchera2019} demonstrated that the purity and statistics of the data are improved with the use of ML.

% From DB (06/19/2020)
From the experimenter's point of view, it is clear that the method used in \cite{Bradt2018} to identify and filter events is not the most efficient computationally. The ML methods explored in \cite{Kuchera2019} are a step forward, but they still rely on supervised learning methods that require data labeling, a time-consuming and error-prone process. The aim of the present study is to investigate unsupervised learning methods that bypass the labeling step, and form classes of events independently from the experimenter's input. The task of labeling the different classes is then much less time-consuming and can potentially lead to the discovery of unforeseen types of events.

@@

In this section we give a brief overview of the data, for a more in-depth consideration we refer the reader to \cite{Mittig2015}, \cite{Suzuki2012} and  \cite{Bradt2017a}.

>>>>

\begin{abstract}
% Modified by DB (6/8/2020)
This article presents novel applications of machine learning methods to the problem of event separation in an active target detector, the Active-Target Time Projection Chamber (AT-TPC) \cite{Bradt2017}. 
The overarching goal is to identify classes of events in the early stages of the data analysis, thereby improving its efficiency by limiting the computationally expensive processing to events of interest only, which are often not the most abundant.
%Machine learning presents an interesting avenue for researchers operating the AT-TPC, as traditional analysis methods of AT-TPC data are computationally expensive and have to fit all tracks against the event type of interest. The latter presents a considerable challenge when either the space of reactions is not known prior to the analysis, or the event type of interest is not the most abundant. 
The application of unsupervised clustering  algorithms to the analysis of two-dimensional projections of particle tracks from a resonant proton scattering experiment on $^{46}$Ar is introduced. We explore the performance of autoencoder neural networks and a pre-trained VGG16 \cite{Simonyan2014} convolutional neural network.
%on two tasks: a semi-supervised classification task and the unsupervised clustering of particle tracks. 
%On the semi-supervised task, we find that a logistic regression classifier trained on small labeled subsets of the latent space of these models perform very well, both on simulated and real data. On the clustering task, 
We find that a $K$-means algorithm applied to the simulated data in the VGG16 latent space forms almost perfect clusters. Additionally, the VGG16+$K$-means approach finds high purity clusters of proton events for real experimental data. We explore also the application of autoencoder neural networks to clustering by implementing. With autoencoder neural networks we find improved descriptions of data from experiments.

@@

The Active-Target Time Projection Chamber (AT-TPC) \cite{Bradt2017} is a novel type of detector designed specifically for nuclear physics experiments where the energies of the recoiling particles are very low compared to the energy required to escape target material. The luminosity of nuclear physics experiments performed with fixed targets is directly proportional to the amount of material encountered by the beam. On the other hand, for several classes of experiments the detection of recoil particles is paramount, therefore limiting the target thickness. In addition, the properties of the recoil particles are modified while traversing the target material, affecting the resolutions that can be achieved. This necessary balance between luminosity and resolution is particularly difficult when performing experiments with rare isotope beams, because of the low intensities available.

The concept of active targets aim at mitigating this compromise, by turning the target itself into a detector \cite{BECEIRONOVO2015}. Most active target detectors such as the AT-TPC are composed of a time projection chamber (TPC) where the detector gas is at the same time the target material. Recoil particles that originate from a nuclear reaction between a beam nucleus and a gas nucleus can be tracked from the vertex of the reaction to their final position inside the active volume of the target. Their properties can therefore be measured without any loss of resolution regardless of the amount of material traversed by the beam. At the same time, the detection efficiency is dramatically increased by the very large solid angle covered in this geometry. A direct consequence of this concept is the inclusiveness of the experimental data recorded by this type of detector: any nuclear reaction happening within the target is recorded. Although this sounds like an advantage from the scientific point of view, it poses great challenges during the analysis phase, that are reminiscent of bubble chamber times and on par with event classification challenges in particle physics today, see for example the recent review of Mehta {\em et al.} \cite{mehta2019}. More often than not, the reaction channel of interest has one of the lowest cross sections. When analyzing the data one is therefore faced with the task of sorting out the corresponding events from the {\em background} of other reaction channels.

% motivating ML work:
Because TPCs produce three-dimensional images of charged particle tracks, the event identification task is often akin to a visual inspection (comparable to the analyses made in the bubble chamber era), which is not practical nowadays because of the large quantities of data. Machine learning techniques then appear as a promising prospect, in particular in the image recognition domain where much progress has been made recently \cite{mehta2019}. In addition, machine learning (ML) algorithms offer new possibilities such as the potential discovery of unforeseen phenomena that would have been missed by more traditional analysis methods. Prior work has demonstrated the ability to apply supervised classification machine learning methods to AT-TPC data when a labeled training set is available, whether through hand-labeled data or labeled simulated data (a {\em transfer learning} application) \cite{Kuchera2019}. In some experiments, a labeled data set is unavailable. This could be  due to the inability to hand-label events, or the case where one does not know {\em a priori} the types and behaviors of the reactions present in the detector in order to generate a labeled and simulated data set. In the latter case, there must also exist a validation data set of real data, still requiring the ability to label a subset of the real data. The unsupervised separation of event types, or {\em clustering}, based on a set of ML algorithms is hereby examined, using experimental data recorded by the AT-TPC during its commissioning experiment from a radioactive $^{46}$Ar beam reacting on an isobutane target composed of proton and carbon nuclei.

@@

The detector was placed inside the bore of an MRI solenoid energized to $\sim 2$ Tesla. This axial magnetic field served the purpose of bending the trajectories of the recoil particles in order to i) increase their length and ii) provide a measurement of their bending radius, directly related to their magnetic rigidity. Because the recoil particles travel in gas, they slow down and eventually stop, therefore their trajectories are described by three-dimensional spirals (see \cite{Bradt2017}). One of the difficulties encountered in the analysis is that the shape of these spirals does not have an analytical form because it follows the energy-loss profile of the particles. It therefore needs to be simulated via an integration, which is numerically costly. Other difficulties are related to several experimental effects that deteriorate the quality of the data, namely saturation and cross-talk effects, as well as random noise.

The method used in \cite{Bradt2018} to analyze the data followed a three-phase sequence: cleaning, filtering and fitting. Traditional methods were used to perform each of these tasks, and ultimately extract the scientific information, but there were severe limitations and high computational costs that become prohibitive in data sets larger by an order of magnitude.  In particular, as for the data presented and analyzed here, with data sets in the terabyte region or larger, these more traditional methods become prohibitively expensive from a computational stand.
% Begin added by DB (06/10/2020)
The cleaning was performed using a combination of linear and circular Hough transforms on a 2-dimensional projection of the tracks \cite{Bradt2017}. The following filtering and fitting phases were performed simultaneously, by defining the cost function to the fitting algorithm as a sum of three $\chi^2$ components based on i) the position of the track in space, ii) the energy deposited on each pixel of the sensor plane, and iii) the location of the vertex of the reaction. While various fitting algorithms were tested, the most accurate was a Monte-Carlo algorithm that explored the six-dimensional phase space of the particle's kinematics parameters, reducing it progressively at each iteration step until the desired accuracy was reached \cite{Bradt2017}. Although this algorithm ended up being the most accurate, it is extremely costly computationally because of the very large number of simulated tracks needed for each event. The filtering was performed by setting limits to the $\chi^2$ distributions, below which the events were assigned as proton scattering. This is a very inefficient method because it requires performing the fitting for all events, including those that are not of interest. Pioneering work on event identification using machine learning methods, namely the use of a pre-trained convolutional neural networks (CNN), later showed that the filtering phase would better be performed using this type of technique \cite{Kuchera2019}. 
% End added by DB (06/10/2020)
In addition, the authors of Ref.~\cite{Kuchera2019} demonstrated that the purity and statistics of the data are improved with the use of machine learning methods.

% From DB (06/19/2020)
From the experimenter's point of view, it is clear that the method used in \cite{Bradt2018} to identify and filter events is not the most efficient computationally. The ML methods explored in \cite{Kuchera2019} are a step forward, but they still rely on supervised learning methods that require data labeling, a time-consuming and error-prone process. The aim of the present study is to investigate unsupervised learning methods that bypass the labeling step, and form classes of events independently from the experimenter's input. The task of labeling the different classes is then much less time-consuming and can potentially lead to the discovery of unforeseen types of events. Furthermore, it allows us to process larger amounts of data in a much shorter time.

@@

In this section we give a brief overview of the data, for a more in-depth consideration we refer the reader to Refs.~\cite{Mittig2015,Suzuki2012,Bradt2017a}.
